Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-11-12 doc/paper-lcr.tex
Edited introduction

Along lines of sight containing strong gravitational lenses, the
perturbative effects of line of sight mass structure have been found to
be particularly important, with foreground and background structures
having a significant effect on the inferred lensing cross-section
\citep[\eg][]{WongEtal2012} and distance ratios
\citep[][]{DalalEtal2005}. Indeed, \citet{SuyuEtal2010} found that in
time delay lens cosmography the so-called ``external convergence''
$\kappa$ due to mass structures along the (unusually over-dense) line
of sight to the quadruply-imaged radio source B1608$+$656 was the
dominant source of uncertainty in their 6\% measurement of Hubble's
constant. The problem is that one simply doesn't know how magnified and
distorted the source has been, because the unlensed source is
unobservable. The external shear can potentially be recovered in the
mass-modelling of the lens, but the convergence is undetectable from the
image positions, shapes and relative fluxes; this is the famous {\emph{ 
mass sheet-degeneracy}} \citep[see e.g.][for details]{FalcoEtal1985}.

How can we correct for this perturbative lensing and make accurate
measurements through the Universe? When inferring global quantities
(such as the cosmological parameters), combining the results from a
large number of independent objects will tend to reduce the uncertainty
due to external convergence, as the ensemble-averaged magnification
approaches zero for an unbiased sample \citep[\eg][]{Holz+Linder2005}.
If the sample is biased, however, some residual systematic error will
remain even as the statistical uncertainty decreases. Lensing bias will
be present (to varying degrees) in any sample {\it selected} by
brightness (which includes magnification), or any other quantity that
depends on $\kappa$. This bias may be introduced either at the object
detection stage, or later on when making decisions about which objects
to study further.  Moreover, we expect these biases to be strongest in
the strong lens systems, where magnification is highest. Large samples
of lenses are expected to be discovered in the next decade in
ground-based optical imaging surveys \citep{Oguri+Marshall2010}: both
the detectability of these lenses,  and the selection of sub-samples of
them for further observation, could be sensitive to the external
convergence.

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected observationally 
by measuring the
small distortions it induces on the images of many background galaxies
in the field \citep[see e.g.][for a review]{Schneider2006}. In a 30''
radius aperture around the strong lens Q0957$+$561,
\citet{NakajimaEtal2009} used deep HST imaging to measure a weak lensing
convergence of $0.17\pm0.06$, which was used in the time delay distance
measurement of \citep{FadelyEtal2009}. While this system has an
unusually high external convergence due to its location on the edge of a
cluster of galaxies, the precision with which the convergence at the
lens position can be estimated is indicative of what can be achieved
with high quality weak lensing shear data. We note that this uncertainty
is not dependent on the value of the convergence, but instead depends
primarily on the number density of weakly lensed, measurable galaxies
available.

@@

The probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$,  is the  key
quantity we seek, since by marginalising over this distribution we can
correctly propagate the uncertainties due to line-of-sight perturbations
(as \citeauthor{FadelyEtal2009} did).  
%
\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gamma|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.

These observational investigations have provided some very useful
insight into the problem of convergence estimation.  Coming from a more
theoretical direction, large numerical simulations have been used to
predict, from first principles, the distribution of likely external
shears and convergences along various lines of sight, including those
containing strong lens systems.  \citet{Holder+Schechter2003} and
\citet{DalalEtal2005} carried out ray-tracing calculations in N-body
simulations to estimate the distribution of external shear values. 
\citet{HilbertEtal2009} performed similar ray-tracing experiments
through the Millenium Simulation \citep{SpringelEtal2005}, generating a
predicted $\kappa$ at every position in a simulated sky.
\citet{HilbertEtal2009} found that \MS lines of sight with strong lenses
were not biased towards high $\kappa$, although selection functions for
discovering or using particular samples of strong lenses were not taken
into account. Addressing the possibility of using weak lensing shear to
reconstruct the magnification~$\mu$ that affects cosmography with
standard candles (and standard sirens), \citet{HilbertEtal2011} used the
\MS ray-tracing results to explore optimal mapping techniques and
quantify their performance. They found that deep weak lensing shear
surveys (with background galaxy number densities of 100 arcmin$^{-2}$)
would enable $\Pr(\mu|\data)$ to be estimated such that the distance
uncertainty per object would be reduced by  20\% on average, and by 40\%
in regions of low convergence, leading to the suggestion that weak
lensing maps be used to identify the (low convergence)  lines of sight
that give higher precision distance estimates.

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy overdensity, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed overdensity towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappa$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappa)$, which was then marginalised over when inferring the time
delay distance in this system. 
% B1608$+$656 is indeed a high $\kappa$
% line of sight, but it does not reside in a massive cluster environment.
In a companion paper to this one, Greene et al (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogs and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{GunnarssonEtal2006}, \citeauthor{JonssonEtal2010}, 
\citeauthor{WongEtal2011} and others, with the
idea of calibrating to simulations from \citeauthor{SuyuEtal2010}, in
order to put the halo model for  estimating the weak lensing convergence
due to mass along the line of sight to a distant source into a sound
framework for evaluating this important source of systematic
uncertainty, as we look forward to ever increasing sample sizes and
statistical precision. Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogs first to
{\it calibrate} the reconstruction, and account for unseen mass and
voids, and then to {\it test} the accuracy of the line of sight mass
reconstruction under this assumption. Probabilistically assigning mass
and redshift to every observed galaxy in a given observed field, we
generate Monte Carlo sample line of sight mass distributions, and so
construct the PDF $\Pr(\kappa|\data)$ for that field. We then emulate
the combination of many such PDFs to quantify the residual biases that
would be translated to the global (including cosmological) parameters.
In doing so, we aim to answer the following questions:

>>>>

Along lines of sight containing strong gravitational lenses, the
perturbative effects of line of sight mass structure have been found to
be particularly important, with foreground and background structures
having a significant effect on the inferred lensing cross-section
\citep[\eg][]{WongEtal2012} and distance ratios
\citep[][]{DalalEtal2005}. Indeed, \citet{SuyuEtal2010} found that in
time delay lens cosmography the so-called ``external convergence''
$\kappa$ due to mass structures along the (unusually over-dense) line
of sight to the quadruply-imaged radio source B1608$+$656 was the
dominant source of uncertainty in their 5\% measurement of the time delay distance. The external shear can potentially be recovered in the
mass-modelling of the lens, but the convergence is undetectable from the
image positions, shapes and relative fluxes; this is the famous {\emph{ 
mass sheet-degeneracy}} \citep[see e.g.][for details]{FalcoEtal1985}.

When inferring global quantities
(such as the cosmological parameters), combining the results from a
large number of independent objects will tend to reduce the uncertainty
due to external convergence \citep[\eg][]{Holz+Linder2005}, unless the lines of sight are not representative of the global population. If the line of sight selection function is not correctly modelled some residual systematic error will
remain even as the statistical uncertainty decreases. A selection function may be introduced either at the object
detection stage, or later on when making decisions about which objects
to study further. Large samples
of lenses are expected to be discovered in the next decade in
ground-based optical imaging surveys \citep{Oguri+Marshall2010}: both
the detectability of these lenses,  and the selection of sub-samples of
them for further observation, could be sensitive to the external
convergence.

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected observationally 
by measuring the
small distortions it induces on the images of many background galaxies
in the field \citep[see e.g.][for a review]{Schneider2006}. 
\citet{NakajimaEtal2009} used deep HST imaging to infer $\kappax=0.17\pm0.06$ from weak lensing measurements; this measurement was used in the time delay distance
measurement of \citet{FadelyEtal2009}. A precise weak lensing estimate of $\kappax$ can only be made if the number density of of weakly lensed, measurable galaxies is high.

@@

In order to propogate the effect of line-of-sight perturbations into 
uncertainties on time-delay distances we require the probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$
% PHIL's REVIEWS OF THE LITRATURE
\comments{\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gamma|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.
}
% TOM's DISTILLING OF SAID REVIEWS.
Previous investigation of line-of-sight perturbations have focused on 
magnification and shear.\citet{GunnarssonEtal2006} used a galaxy halo model combined with simple scaling relations to reconstruct
the line of sight mass distribution and found that the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two. Conversely \citet{KarpenkaEtal2012} used the brightness dispersion in type Ia supernovae to infer the parameters of both their galaxy halo model and mass-to-light scaling relation. Both \citeauthor{KarpenkaEtal2012} and \citet{JonssonEtal2010} were able to 
detect lensing effects in the supernova data this way.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data,
\citep{WilliamsEtal2006;MomchevaEtal2006} of galaxies close to the line
of sight. Comparing the predicted shear
distributions with the external shear demanded by the strong lens model 
they found significant
discrepancies, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies. \citeauthor{WongEtal2011} also found that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear.

Large numerical simulations have been used to
predict, from first principles, the distribution of likely external
shears and convergences along various lines of sight, including those
containing strong lens systems.  \citet{Holder+Schechter2003} and
\citet{DalalEtal2005} carried out ray-tracing calculations in N-body
simulations to estimate the distribution of external shear values. 
\citet{HilbertEtal2009} performed similar ray-tracing experiments
through the Millenium Simulation \citep{SpringelEtal2005}, generating a
predicted $\kappa$ at every position in a simulated sky.
\citet{HilbertEtal2009} found that \MS lines of sight with strong lenses
were not biased towards high $\kappa$, although selection functions for
discovering or using particular samples of strong lenses were not taken
into account.

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy overdensity, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed overdensity towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappa$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappa)$, which was then marginalised over when inferring the time
delay distance in this system. A similar proceedure was used in \citet{SuyuEtal2012b}.
% B1608$+$656 is indeed a high $\kappa$
% line of sight, but it does not reside in a massive cluster environment.
In a companion paper to this one, Greene et al. (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogs and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$. For the most overdense lines of sight the
Green et al. results  show a $\sim$20\% improvement over
number counts alone, but little improvement for less dense lines.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{GunnarssonEtal2006}, \citeauthor{JonssonEtal2010}, 
\citeauthor{WongEtal2011} and others, with the
idea of calibrating to simulations from \citeauthor{SuyuEtal2010}.
Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogs first to
{\it calibrate} the reconstruction, and account for unseen mass and
voids, and then to {\it test} the accuracy of the line of sight mass
reconstruction under this assumption. Probabilistically assigning mass
and redshift to every observed galaxy in a given observed field, we
generate Monte Carlo sample line of sight mass distributions, and so
construct the PDF $\Pr(\kappa|\data)$ for that field. We then emulate
the combination of many such PDFs to quantify the residual biases that
would be translated to the global (including cosmological) parameters.
In doing so, we aim to answer the following questions:
